---
title: "Bayesian G-Study Base Model Without Task-Harness"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  rows_csv: "hal_binary_rows.csv"
  chains: 4
  iter: 2000
  warmup: 1000
  adapt_delta: 0.95
  max_treedepth: 12
  seed: 20260527
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required <- c("brms", "posterior", "dplyr", "readr", "tibble", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing, collapse = ", "),
    "\nInstall them with install.packages(...) before knitting this file."
  )
}

library(brms)
library(posterior)
library(dplyr)
library(readr)
library(tibble)
library(ggplot2)

stan_backend <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"
```

## Goal

This notebook fits the Bayesian G-study model \emph{without} the
`task_harness` random effect:

```r
correct ~ 1 +
  (1 | benchmark) +
  (1 | task_nested) +
  (1 | model_name) +
  (1 | harness_name) +
  (1 | model_harness) +
  (1 | benchmark_harness)
```

This answers: how much outcome structure is already accounted for by benchmark,
item difficulty, model, harness, model--harness compatibility, and
benchmark-specific harness effects? We then inspect residuals grouped by
`task_harness` as evidence for the remaining item-specific harness component.

## Data

```{r data}
rows <- read_csv(params$rows_csv, show_col_types = FALSE) |>
  mutate(
    correct = as.integer(correct),
    benchmark = factor(benchmark),
    task_nested = factor(task_nested),
    model_name = factor(model_name),
    harness_name = factor(scaffold_name),
    model_harness = interaction(model_name, harness_name, drop = TRUE, sep = " | "),
    benchmark_harness = interaction(benchmark, harness_name, drop = TRUE, sep = " | "),
    task_harness = interaction(task_nested, harness_name, drop = TRUE, sep = " | ")
  )

tibble(
  n_rows = nrow(rows),
  mean_accuracy = mean(rows$correct),
  benchmarks = n_distinct(rows$benchmark),
  tasks = n_distinct(rows$task_nested),
  models = n_distinct(rows$model_name),
  harnesses = n_distinct(rows$harness_name),
  model_harness_pairs = n_distinct(rows$model_harness),
  benchmark_harness_pairs = n_distinct(rows$benchmark_harness),
  task_harness_pairs = n_distinct(rows$task_harness)
)
```

## Base Bayesian G-Study Model

```{r fit-base}
priors <- c(
  prior(normal(0, 2.5), class = "Intercept"),
  prior(exponential(1), class = "sd")
)

fit_base <- brm(
  correct ~ 1 +
    (1 | benchmark) +
    (1 | task_nested) +
    (1 | model_name) +
    (1 | harness_name) +
    (1 | model_harness) +
    (1 | benchmark_harness),
  data = rows,
  family = bernoulli(link = "logit"),
  prior = priors,
  chains = params$chains,
  iter = params$iter,
  warmup = params$warmup,
  seed = params$seed,
  cores = min(params$chains, parallel::detectCores()),
  control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth),
  backend = stan_backend,
  file = "brms_bayesian_gstudy_base_fit",
  file_refit = "on_change"
)

fit_base
```

## Variance Components Already Accounted For

For Bernoulli-logit models, residual variance is fixed on the latent scale at
`pi^2 / 3`. We report variance components and ICC-like variance shares.

```{r variance-components}
draws <- as_draws_df(fit_base)
sd_cols <- grep("^sd_", names(draws), value = TRUE)

variance_draws <- draws |>
  select(all_of(sd_cols)) |>
  mutate(across(everything(), ~ .x^2)) |>
  rename_with(~ sub("^sd_", "var_", .x))

residual_var <- pi^2 / 3
total_var <- rowSums(variance_draws) + residual_var

share_draws <- variance_draws |>
  mutate(across(everything(), ~ .x / total_var)) |>
  mutate(var_residual_logit = residual_var / total_var)

variance_summary <- bind_rows(
  summarise_draws(variance_draws, mean, median, ~ quantile(.x, 0.05), ~ quantile(.x, 0.95)) |>
    as_tibble() |>
    rename(component = variable, q05 = `5%`, q95 = `95%`) |>
    mutate(scale = "variance"),
  summarise_draws(share_draws, mean, median, ~ quantile(.x, 0.05), ~ quantile(.x, 0.95)) |>
    as_tibble() |>
    rename(component = variable, q05 = `5%`, q95 = `95%`) |>
    mutate(scale = "variance_share")
) |>
  mutate(component = gsub("^var_", "", component))

variance_summary

write_csv(variance_summary, "brms_gstudy_base_variance_components.csv")
```

```{r variance-plot, fig.width=7, fig.height=4}
variance_summary |>
  filter(scale == "variance_share") |>
  mutate(component = reorder(component, mean)) |>
  ggplot(aes(x = component, y = mean)) +
  geom_col(fill = "#3b6fb6") +
  geom_errorbar(aes(ymin = q05, ymax = q95), width = 0.2) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Posterior mean variance share",
    title = "Base Bayesian G-study variance shares"
  ) +
  theme_minimal(base_size = 11)

ggsave("brms_gstudy_base_variance_shares.png", width = 7, height = 4, dpi = 200)
```

## Residual Task-Harness Structure

This section estimates how much task-harness structure remains after the base
model has accounted for benchmark, task, model, harness, model-harness, and
benchmark-harness terms. It is not a substitute for the full random-effect
`task_harness` model, but it is a fast diagnostic for whether that term is
worth fitting.

```{r residual-task-harness}
pred <- fitted(fit_base, summary = TRUE)[, "Estimate"]

residual_rows <- rows |>
  mutate(
    predicted = pred,
    response_residual = correct - predicted
  )

task_harness_residuals <- residual_rows |>
  group_by(task_harness, task_nested, harness_name) |>
  summarise(
    n = n(),
    accuracy = mean(correct),
    mean_predicted = mean(predicted),
    mean_residual = mean(response_residual),
    abs_mean_residual = abs(mean_residual),
    .groups = "drop"
  ) |>
  arrange(desc(abs_mean_residual), desc(n))

residual_summary <- tibble(
  n_task_harness_cells = nrow(task_harness_residuals),
  unweighted_var_cell_residual_mean = var(task_harness_residuals$mean_residual),
  weighted_var_cell_residual_mean = weighted.mean(
    (task_harness_residuals$mean_residual - weighted.mean(task_harness_residuals$mean_residual, task_harness_residuals$n))^2,
    task_harness_residuals$n
  ),
  mean_abs_cell_residual = mean(abs(task_harness_residuals$mean_residual)),
  weighted_mean_abs_cell_residual = weighted.mean(abs(task_harness_residuals$mean_residual), task_harness_residuals$n)
)

residual_summary
task_harness_residuals |> filter(n >= 5) |> slice_head(n = 20)

write_csv(residual_summary, "brms_gstudy_base_task_harness_residual_summary.csv")
write_csv(task_harness_residuals, "brms_gstudy_base_task_harness_residuals.csv")
```

## Interpretation

The base variance-component table shows what the model has already accounted
for before introducing item-specific harness effects. The grouped residual
diagnostic then asks whether the remaining errors cluster by `task_harness`.
If the residual summary is large, or if many task-harness cells show systematic
over- or under-performance, that motivates the full model with:

```r
(1 | task_harness)
```

as the next robustness check.

