---
title: "Bayesian G-Study for HAL Agent Benchmark Outcomes"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  rows_csv: "hal_binary_rows.csv"
  chains: 4
  iter: 2000
  warmup: 1000
  adapt_delta: 0.95
  max_treedepth: 12
  seed: 20260527
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required <- c("brms", "posterior", "dplyr", "readr", "tibble", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing, collapse = ", "),
    "\nInstall them with install.packages(...) before knitting this file."
  )
}

library(brms)
library(posterior)
library(dplyr)
library(readr)
library(tibble)
library(ggplot2)

stan_backend <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"
```

## Goal

This notebook fits a Bayesian generalizability-style model to the HAL binary
response rows. The question is where outcome variation lives before we fit the
IRT model: benchmark, task, model, harness, model--harness compatibility, or
task--harness interaction.

Because the response matrix is sparse and unbalanced, this is not a classical
balanced random-effects G-study. It is a Bernoulli-logit crossed random-effects
model with partial pooling.

## Data

```{r data}
rows <- read_csv(params$rows_csv, show_col_types = FALSE) |>
  mutate(
    correct = as.integer(correct),
    benchmark = factor(benchmark),
    task_nested = factor(task_nested),
    model_name = factor(model_name),
    harness_name = factor(scaffold_name),
    model_harness = interaction(model_name, harness_name, drop = TRUE, sep = " | "),
    benchmark_harness = interaction(benchmark, harness_name, drop = TRUE, sep = " | "),
    task_harness = interaction(task_nested, harness_name, drop = TRUE, sep = " | ")
  )

tibble(
  n_rows = nrow(rows),
  mean_accuracy = mean(rows$correct),
  benchmarks = n_distinct(rows$benchmark),
  tasks = n_distinct(rows$task_nested),
  models = n_distinct(rows$model_name),
  harnesses = n_distinct(rows$harness_name),
  model_harness_pairs = n_distinct(rows$model_harness),
  task_harness_pairs = n_distinct(rows$task_harness)
)
```

## Model

The reportable model is:

```r
correct ~ 1 +
  (1 | benchmark) +
  (1 | task_nested) +
  (1 | model_name) +
  (1 | harness_name) +
  (1 | model_harness) +
  (1 | benchmark_harness) +
  (1 | task_harness)
```

Each term is a random intercept. For example, `(1 | model_name)` estimates a
partially pooled model-level ability offset, while `(1 | model_harness)`
estimates whether a particular model--harness combination is unusually strong or
weak after accounting for the model and harness main effects. The
`task_harness` term asks whether a harness is unusually strong or weak on
particular items after accounting for task difficulty and average harness
strength.

For Bernoulli-logit models, residual variance is not freely estimated as in a
Gaussian model. On the latent logit scale, the conventional residual variance is
fixed at `pi^2 / 3`. We therefore report posterior variance components and
ICC-like shares:

```text
Var(facet) / (sum Var(facets) + pi^2 / 3)
```

```{r fit}
priors <- c(
  prior(normal(0, 2.5), class = "Intercept"),
  prior(exponential(1), class = "sd")
)

fit <- brm(
  correct ~ 1 +
    (1 | benchmark) +
    (1 | task_nested) +
    (1 | model_name) +
    (1 | harness_name) +
    (1 | model_harness) +
    (1 | benchmark_harness) +
    (1 | task_harness),
  data = rows,
  family = bernoulli(link = "logit"),
  prior = priors,
  chains = params$chains,
  iter = params$iter,
  warmup = params$warmup,
  seed = params$seed,
  cores = min(params$chains, parallel::detectCores()),
  control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth),
  backend = stan_backend,
  file = "brms_bayesian_gstudy_fit",
  file_refit = "on_change"
)

fit
```

## Variance Components

```{r variance-components}
draws <- as_draws_df(fit)

sd_cols <- grep("^sd_", names(draws), value = TRUE)

variance_draws <- draws |>
  select(all_of(sd_cols)) |>
  mutate(across(everything(), ~ .x^2)) |>
  rename_with(~ sub("^sd_", "var_", .x))

residual_var <- pi^2 / 3
total_var <- rowSums(variance_draws) + residual_var

share_draws <- variance_draws |>
  mutate(across(everything(), ~ .x / total_var)) |>
  mutate(var_residual_logit = residual_var / total_var)

variance_summary <- bind_rows(
  summarise_draws(variance_draws, mean, median, ~ quantile(.x, 0.05), ~ quantile(.x, 0.95)) |>
    as_tibble() |>
    rename(component = variable, q05 = `5%`, q95 = `95%`) |>
    mutate(scale = "variance"),
  summarise_draws(share_draws, mean, median, ~ quantile(.x, 0.05), ~ quantile(.x, 0.95)) |>
    as_tibble() |>
    rename(component = variable, q05 = `5%`, q95 = `95%`) |>
    mutate(scale = "variance_share")
) |>
  mutate(component = gsub("^var_", "", component))

variance_summary

write_csv(variance_summary, "brms_gstudy_variance_components.csv")
```

```{r variance-plot, fig.width=7, fig.height=4}
variance_summary |>
  filter(scale == "variance_share") |>
  mutate(component = reorder(component, mean)) |>
  ggplot(aes(x = component, y = mean)) +
  geom_col(fill = "#3b6fb6") +
  geom_errorbar(aes(ymin = q05, ymax = q95), width = 0.2) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Posterior mean variance share",
    title = "Bayesian G-study variance shares on the logit scale"
  ) +
  theme_minimal(base_size = 11)

ggsave("brms_gstudy_variance_shares.png", width = 7, height = 4, dpi = 200)
```

## Optional Model Comparisons

These lighter models are useful if the full `task_harness` model is slow or if
we want to report the incremental contribution of interaction terms.

```{r optional-comparisons, eval=FALSE}
fit_main <- brm(
  correct ~ 1 +
    (1 | benchmark) +
    (1 | task_nested) +
    (1 | model_name) +
    (1 | harness_name),
  data = rows,
  family = bernoulli(link = "logit"),
  prior = priors,
  chains = params$chains,
  iter = params$iter,
  warmup = params$warmup,
  seed = params$seed,
  cores = min(params$chains, parallel::detectCores()),
  control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth),
  backend = stan_backend,
  file = "brms_bayesian_gstudy_main_fit",
  file_refit = "on_change"
)

fit_model_harness <- update(
  fit_main,
  formula = correct ~ 1 +
    (1 | benchmark) +
    (1 | task_nested) +
    (1 | model_name) +
    (1 | harness_name) +
    (1 | model_harness),
  file = "brms_bayesian_gstudy_model_harness_fit",
  file_refit = "on_change"
)

loo(fit_main, fit_model_harness, fit)
```

## Paper Language Template

If the posterior variance shares show substantial mass on `task_nested`,
`model_harness`, or `task_harness`, the paper can summarize the finding as:

> A Bayesian G-study-style Bernoulli mixed model indicates that benchmark labels
> are too coarse for agent measurement: substantial variance is attributable to
> item identity and to interactions involving harnesses. This motivates the IRT
> model's item-level difficulty and discrimination parameters, and motivates a
> residual harness-feature analysis asking which tools help which models on
> which item types.
